---
title: "Spatial Viz"
number-sections: false
---

Use this file for practice with the **spatial viz** in-class activity.  Refer to the class website for details.

# Review
```{r}
# Load tidyverse package for plotting and wrangling
library(tidyverse)

# Import the data and only keep 2 addresses
energy <- read.csv("https://mac-stat.github.io/data/MacNaturalGas.csv") |> 
  mutate(date = as.Date(paste0(month, "/1/", year), "%m/%d/%Y")) |> 
  filter(address != "c")

# Check it out
head(energy)
```
- Construct a plot that addresses each research question
- Include a 1-sentence summary of the plot.

## Example 1
What was the range and typical energy used each month, as measured by `therms`? How does this differ by address?
```{r}
ggplot(energy, aes(x = therms, color = address)) +
  geom_density()
```
The range of energy use for both addresses goes from 0 to about 450. Address B has a higher peak than address A, but there generally seem to be the most months that used 0-100 therms of energy, with a secondary peak at about 350-400 therms.

## Example 2
How did energy use (`therms`) change over time (`date`) at the two addresses?

# New Stuff

- Types of spatial visualization:
  - Point maps: plotting locations of individual observations
  - Contour maps: plotting the distribution of observations
  - Chloropleth maps: plotting outcomes in different regions
- Can be **static** or **dynamic**

# Exercises

- There are two pieces to every map:
  - A dataset (**location coordinates** for point maps or **variable coordinates** for chloropleth maps)
  - A background map 
- `leaflet` can be used to create interactive maps

## Exercise 1
```{r}
fave_places <- read.csv("https://hash-mac.github.io/stat112site-s25/data/our_fave_places.csv")

# Check it out
head(fave_places)
```
### Part a
```{r}
# Load the leaflet package
library(leaflet)

# Just a plotting frame
leaflet(data = fave_places)
```
```{r}
# Now what do we have?
# this adds the base map
leaflet(data = fave_places) |> 
  addTiles()
```
```{r}
# Now what do we have?
# longitude and latitude refer to the variables in our data
# this loads the datapoints
leaflet(data = fave_places) |> 
  addTiles() |> 
  addMarkers(lng = ~longitude, lat = ~latitude)
```
```{r}
# Since we named them "longitude" and "latitude", the function
# automatically recognizes these variables. No need to write them!
leaflet(data = fave_places) |> 
  addTiles() |> 
  addMarkers()
```
### Part b
For most places, it's possible to zoom in to street-level detail! Some places have more detail than others.

## Exercise 2
```{r}
# Load package needed to change color
library(gplots)

# We can add colored circles instead of markers at each location
leaflet(data = fave_places) |> 
  addTiles() |> 
  addCircles(color = col2hex("red"))
```
```{r}
# We can change the background
# Mark locations with yellow dots
# And connect the dots, in their order in the dataset, with green lines
# (These green lines don't mean anything here, but would if this were somebody's travel path!)
leaflet(data = fave_places) |>
  addProviderTiles("USGS") |>
  addCircles(weight = 10, opacity = 1, color = col2hex("yellow")) |>
  addPolylines(
    lng = ~longitude,
    lat = ~latitude,
    color = col2hex("green")
  )
```

- `addProviderTiles()` change the base map.
  - `providers` in the console - list of options
- `addMarkers()` and `addCircles()` mark locations.
  - `?addControl`
  - `weight` = thickness of the lines/points/pixels
  - `opacity` = transparency, like `alpha`
  - color needs to be in hex form. use `col2hex()`
  
## Exercise 3
```{r}
# Import starbucks location data
starbucks <- read.csv("https://mac-stat.github.io/data/starbucks.csv")
```
```{r}
# MN only
# Don't worry about the syntax
starbucks_mn <- starbucks |>   
  filter(Country == "US", State.Province == "MN")
```
```{r}
# Leaflet map of MN starbucks
leaflet(data = starbucks_mn) |>
  addTiles() |>
  addMarkers()
```

## Exercise 4
```{r}
ggplot(starbucks, aes(x = Latitude, y = Longitude)) +
  geom_point(alpha = 0.2, size = 0.2)
```
- There's no background map, so it's hard to visualize where the locations are if you don't know what the data is for.

## Exercise 5
### Part a
```{r}
# Load the package
library(rnaturalearth)

# Get info about country boundaries across the world
# in a "sf" or simple feature format
world_boundaries <- ne_countries(returnclass = "sf")
```

### Part b
```{r}
# What does this code produce?
# What geom are we using for the point map?
# This produces a world map with country boundaries; the geom is "sf" = simple feature
ggplot(world_boundaries) + 
  geom_sf()
```
```{r}
# Load package needed to change map theme
library(mosaic)

# Add a point for each Starbucks
# NOTE: The Starbucks info is in our starbucks data, not world_boundaries
# How does this change how we use geom_point?!
ggplot(world_boundaries) + 
  geom_sf() + 
  geom_point(
    data = starbucks,
    aes(x = Longitude, y = Latitude),
    alpha = 0.3, size = 0.2, color = "darkgreen"
  ) +
  theme_map()
```
### Part c

- The vast majority of Starbucks locations are in America, followed by Europe and East Asia.

## Exercise 6
Instead of world_boundaries <- ne_countries(returnclass = 'sf') we could zoom in on…

- the continent of Africa: `ne_countries(continent = 'Africa', returnclass = 'sf')`
- a set of countries: `ne_countries(country = c('france', 'united kingdom', 'germany'), returnclass = 'sf')`
- boundaries within a country: `ne_states(country = 'united states of america', returnclass = 'sf')`

Part a:
```{r}
# Data for starbucks in Canada, Mexico, and the US
# We'll learn this syntax soon! Don't worry about it now.
starbucks_cma <- starbucks |> 
  filter(Country %in% c('CA', 'MX', 'US'))
```
```{r}
cma_boundaries <- ne_states(
  country = c("canada", "mexico", "united states of america"),
  returnclass = "sf")
```

Part b:
```{r}
# Making the map
# Just the boundaries
ggplot(cma_boundaries) + 
  geom_sf()
```
```{r}
# Add the points
# And zoom in
ggplot(cma_boundaries) + 
  geom_sf() + 
  geom_point(
    data = starbucks_cma,
    aes(x = Longitude, y = Latitude),
    alpha = 0.3,
    size = 0.2,
    color = "darkgreen"
  ) +
  coord_sf(xlim = c(-179.14, -50)) +
  theme_map()
```
## Exercise 7
Zoom in more...Minnesota, Wisconsin, North Dakota, South Dakota

### Part a:
```{r}
# Data for Starbucks only in these states
starbucks_midwest <- starbucks |> 
  filter(State.Province %in% c("MN", "ND", "SD", "WI"))
```
```{r}
# Load packages
library(sf)
library(maps)

# Get the boundaries
midwest_boundaries <- st_as_sf(
  maps::map("county",
            region = c("minnesota", "wisconsin", "north dakota", "south dakota"), 
            fill = TRUE, plot = FALSE))

# Check it out
head(midwest_boundaries)
```
### Part b:
```{r}
ggplot(midwest_boundaries) + 
  geom_sf() + 
  geom_point(
    data = starbucks_midwest,
    aes(x = Longitude, y = Latitude),
    alpha = 0.7,
    size = 0.2, 
    color = 'darkgreen'
    ) + 
  theme_map()
```
## Exercise 8
When there are many overlapping points, it can be tough to visualize areas of higher density. Here is the map of Canada, Mexico, and the US, from Exercise 6:
```{r}
# Point map (we made this earlier)
ggplot(cma_boundaries) + 
  geom_sf() + 
  geom_point(
    data = starbucks_cma,
    aes(x = Longitude, y = Latitude),
    alpha = 0.3,
    size = 0.2,
    color = "darkgreen"
  ) +
  coord_sf(xlim = c(-179.14, -50), ylim = c(14.54, 83.11)) +
  theme_map()
```
Now check out the Contour Map:
```{r}
# What changed in the plot?
# Instead of displaying points, it now displays connected lines, like a contour height map.
# What changed in our code?!
# Instead of using geom_point(), we're now using geom_density_2d()
ggplot(cma_boundaries) + 
  geom_sf() + 
  geom_density_2d(
    data = starbucks_cma,
    aes(x = Longitude, y = Latitude),
    size = 0.2,
    color = "darkgreen"
  ) +
  coord_sf(xlim = c(-179.14, -50), ylim = c(14.54, 83.11)) +
  theme_map()
```
## Exercise 9
Non-spatial data:
```{r}
# State and county level data on elections
elections_by_state <-  read.csv("https://mac-stat.github.io/data/election_2020_by_state.csv")
elections_by_counties <- read.csv("https://mac-stat.github.io/data/election_2020_county.csv")
```
Wrangle this data to present what we're interested in: overall outcome by region, rather than geographic location of observations.
```{r}
# Don't worry about the code!

elections_by_state <- elections_by_state |> 
  filter(state_abbr != "DC") |> 
  select(state_name, state_abbr, repub_pct_20) |> 
  mutate(repub_20_categories = 
           cut(repub_pct_20, 
               breaks = seq(30, 70, by = 5), 
               labels = c("30-34", "35-39", "40-44", "45-49",
                          "50-54", "55-59", "60-64", "65-70"), 
               include.lowest = TRUE))

elections_by_counties <- elections_by_counties |> 
  select(state_name, state_abbr, county_name, county_fips,
          repub_pct_20, median_age, median_rent) |> 
  mutate(repub_20_categories = 
           cut(repub_pct_20, 
               breaks = seq(0, 100, by = 10),
               labels = c("0-9", "10-19", "20-29", "30-39", "40-49",
                          "50-59", "60-69", "70-79", "80-89", "90-100"),
               include.lowest = TRUE))
```

### Part a:
To map the 2020 Republican support in each state (`repub_pct_20`), we need two pieces of information:

- Data on elections in each state (`elections_by_state`)
- A background map of state boundaries in the US
```{r}
# Get the latitude and longitude coordinates of state boundaries
states_map <- map_data("state")

# Check it out
head(states_map)
```
### Part b:
```{r}
# Note where the dataset, elections_by_state, is used
# Note where the background map, states_map, is used
ggplot(elections_by_state, aes(map_id = state_name, fill = repub_pct_20)) +
  geom_map(map = states_map) +
  expand_limits(x = states_map$long, y = states_map$lat) +
  theme_map() 
```
```{r}
# Make it nicer!
ggplot(elections_by_state, aes(map_id = state_name, fill = repub_pct_20)) +
  geom_map(map = states_map) +
  expand_limits(x = states_map$long, y = states_map$lat) +
  theme_map() + 
  scale_fill_gradientn(name = "% Republican", colors = c("blue", "purple", "red"), values = scales::rescale(seq(0, 100, by = 5)))
```
```{r}
# plotting the discretized version, repub_20_categories
ggplot(elections_by_state, aes(map_id = state_name, fill = repub_20_categories)) +
  geom_map(map = states_map) +
  expand_limits(x = states_map$long, y = states_map$lat) +
  theme_map()
```
```{r}
# Load package needed for refining color palette
library(RColorBrewer)

# Now fix the colors
ggplot(elections_by_state, aes(map_id = state_name, fill = repub_20_categories)) +
  geom_map(map = states_map) +
  expand_limits(x = states_map$long, y = states_map$lat) +
  theme_map() + 
  scale_fill_manual(values = rev(brewer.pal(8, "RdBu")), name = "% Republican")
```
### Part c:
```{r}
# Get only the starbucks data from the US
starbucks_us <- starbucks |> 
  filter(Country == "US")

# Map it
ggplot(elections_by_state, aes(map_id = state_name, fill = repub_20_categories)) +
  geom_map(map = states_map) +
  geom_point(
    data = starbucks_us,
    aes(x = Longitude, y = Latitude),
    size = 0.05,
    alpha = 0.2,
    inherit.aes = FALSE
  ) +
  expand_limits(x = states_map$long, y = states_map$lat) +
  theme_map() + 
  scale_fill_manual(values = rev(brewer.pal(8, "RdBu")), name = "% Republican")
```
## Exercise 10
### Part a:
```{r}
# Get the latitude and longitude coordinates of county boundaries
library(socviz)
data(county_map) 

# Check it out
head(county_map)
```
```{r}
# Add 0's at the beginning of any fips_code that's fewer than 5 numbers long
# Don't worry about the syntax
elections_by_counties <- elections_by_counties |> 
  mutate(county_fips = as.character(county_fips)) |> 
  mutate(county_fips = 
           ifelse(nchar(county_fips) == 4, paste0("0", county_fips), county_fips))
```

### Part b:
```{r}
# Using discretized variable 'repub_20_categories' and color scale
ggplot(elections_by_counties, aes(map_id = county_fips, fill = repub_20_categories)) +
  geom_map(map = county_map) +
  scale_fill_manual(values = rev(brewer.pal(10, "RdBu")), name = "% Republican") +
  expand_limits(x = county_map$long, y = county_map$lat) +
  theme_map() +
  theme(legend.position = "right") + 
  coord_equal()
```
## Exercise 11
```{r}
# County-level map of 'median_rent'
ggplot(elections_by_counties, aes(map_id = county_fips, fill = median_rent)) +
  geom_map(map = county_map) +
  scale_fill_gradientn(name = "Rent ($)", colors = c("lightgreen", "darkgreen", "black"), values = scales::rescale(seq(0, 100, by = 5))) +
  expand_limits(x = county_map$long, y = county_map$lat) +
  theme_map() +
  theme(legend.position = "right") + 
  coord_equal()
```
```{r}
# County-level map of 'median_age'
ggplot(elections_by_counties, aes(map_id = county_fips, fill = median_age)) +
  geom_map(map = county_map) +
  scale_fill_gradientn(name = "Age", colors = c("white", "lightblue", "pink"), values = scales::rescale(seq(0, 100, by = 5))) +
  expand_limits(x = county_map$long, y = county_map$lat) +
  theme_map() +
  theme(legend.position = "right") + 
  coord_equal()
```



