---
title: "Reshaping"
number-sections: false
---

Use this file for practice with the **reshaping** in-class activity.  Refer to the class website for details.

# Review
## Example 1
```{r}
library(tidyverse)
penguins <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-07-28/penguins.csv')
```


## Example 2
```{r}
# Using count()


# Using group_by() and summarize()
```

```{r}
penguin_avg <- penguins |> 
  group_by(species, sex) |> 
  summarize(avg_body_mass = mean(body_mass_g, na.rm = TRUE)) |> 
  na.omit()
```

## Example 3
```{r}
# Units of observation = 6
head(penguins)
```
```{r}
# Units of observation = 6
head(penguin_avg)
```
- It doesn't seem that any data is lost; it is just changed into a different format.

# Reshaping Data
Types of data reshaping:

- Aggregate data
  - `groupby()` + `summarize()` gives aggregate information, but loses individual data
- Raw data (reshaped)
  - Retains all information

## Example 4
```{r}
penguin_avg
# make "male" and "female" their own columns
```


## Example 5
```{r}
penguin_avg |> 
pivot_wider(names_from = sex, values_from = avg_body_mass)
```
- units of observation: 3
- we did not lose any information.
- on average, females have less body size than males.

## Example 6
```{r}
penguin_avg_wide <- penguin_avg |> 
  pivot_wider(names_from = sex, values_from = avg_body_mass)

penguin_avg_wide
```
```{r}
# We can either communicate which variables we WANT to collect into a single column (female, male)
penguin_avg_wide |> 
  pivot_longer(cols = c(female, male), names_to = "sex", values_to = "avg_body_mass")
```
```{r}
# Or which variable(s) we do NOT want to collect into a single column (species)
penguin_avg_wide |> 
  pivot_longer(cols = -species, names_to = "sex", values_to = "avg_body_mass")
```

- units of observation: 6
- we did not lose any information.
- here, the variables are in parenthesis because its new input information (naming the new columns), while with `pivot_wider()` they're variables that already exist within the table and are being called on.

## Example 7
```{r}
food <- data.frame(
  customer = rep(c("A", "B"), each = 3),
  restaurant = rep(c("Shish", "FrenchMeadow", "DunnBros"), 2),
  order = c("falafel", "salad", "coffee", "baklava", "pastry", "tea")
)
food
```
```{r}
# Wrangle this data so that the units of observation are customers, spreading the restaurants into separate columns.
food |>
  pivot_wider(names_from = restaurant, values_from = order)
```


```{r}
more_food <- data.frame(
  customer = c("C", "D"),
  Shish = c("coffee", "maza"),
  FrenchMeadow = c("soup", "sandwich"),
  DunnBros = c("cookie", "coffee")
)
more_food
```
```{r}
# Wrangle this data so that the 3 restaurant columns are combined into 1, hence the units of observation are customer / restaurant combinations.
more_food |>
  pivot_longer(cols = -customer, names_to = "restaurant", values_to = "order")
```

# Exercises
## Exercise 1
```{r}
sleep_wide <- read.csv("https://mac-stat.github.io/data/sleep_wide.csv")

head(sleep_wide)
```
### Part a:

- Units of observation: 18

### Part b:
Plot each subject’s reaction time (y-axis) vs the number of days of sleep restriction (x-axis).
Plotting frame:

- ggplot(sleep, aes(y = reaction_time, x = days, color = subject))

Therefore, the table needs to be made longer so that all the days are in the same column.

## Exercise 2
### Part a:
```{r}
sleep_wide |>
  pivot_longer(cols = -Subject, names_to = "days", values_to = "reaction_time") |>
  head(3)
```
### Part b:
```{r}
sleep_long <- sleep_wide |>
  pivot_longer(cols = -Subject,
               names_to = "day",
               names_prefix = "day_",
               values_to = "reaction_time")

head(sleep_long)
```
- This code added "names_prefix", which removes the prefix to the variables - in this case "day_".
- The values in the "day" column are just numerical now.

### Part c:
```{r}
ggplot(sleep_long, aes(y = reaction_time, x = day, color = Subject)) +
  geom_line()
```
- "Subject" is a quantitative variable - we want it to be categorical
- "Day" is a categorical variable - we want it to be quantitative

## Exercise 3
```{r}
sleep_long <- sleep_wide |>
  pivot_longer(cols = -Subject,
               names_to = "day",
               names_prefix = "day_",
               values_to = "reaction_time") |> 
  mutate(Subject = as.factor(Subject), day = as.numeric(day))

# Check it out
# Same data, different class
head(sleep_long)
```
### Part a:
```{r}
# Make a line plot of reaction time by day for each subject
# Put these all on the same frame
ggplot(sleep_long, aes(y = reaction_time, x = day, color = Subject)) +
  geom_line()
```
```{r}
# Make a line plot of reaction time by day for each subject
# Put these all on separate frames (one per subject)
ggplot(sleep_long, aes(y = reaction_time, x = day, color = Subject)) +
  geom_line() +
  facet_wrap(facets = ~Subject)
```

### Part b:

- Generally, it seems that the less sleep a person has, the higher their reaction time is.
- There are some exceptions (332 and 335), and there's variance between each person's reaction time.

## Exercise 4
### Part a:
```{r}
sleep_long |>
   pivot_wider(names_from = day, values_from = reaction_time) |> 
   head()
```
- Because we removed the "day_" prefix, the column labels are just numbers now, which can be confusing.

### Part b:
```{r}
sleep_long |>
  pivot_wider(names_from = day, values_from = reaction_time, names_prefix = "day_") |> 
  head()
```

## Exercise 5
```{r}
# Load data
library(tidyr)
data("billboard")

# Check it out
head(billboard)
```
### Part a:


